---
title: "mapping_lynx_WGS_samples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## Mapping Lynx WGS Data ##

library(sp) #-- need
library(r2dRue)
library(adehabitatHR)
library(rgdal) #-- need
library(ade4)
library(deldir)
library(adehabitatMA)
library(adehabitatLT)
library(shapefiles)
library(spatstat)
library(rgeos)
#library(gpclib)
library(maptools)
library(plyr)
require(ggmap)
require(rgeos)

setwd()
source('F:/work/movement/biostats.R')
```
# Plotting the outline of the US/Canada, provinces, and in particular the St. Lawrence River
```{r}
library(maps)
library(mapproj)
library(mapdata)
library(rgeos)
library(maptools)
library(sp)
library(raster)
library(rgdal)

can0<-getData('GADM', country="CAN", level=0) # Canada
can1<-getData('GADM', country="CAN", level=1) # provinces
# can2<-getData('GADM', country="CAN", level=2) # counties
  
plot(can1)    
spplot(can1, "NAME_1") # colors the provinces and provides
                       # a color-coded legend for them
can1$NAME_1            # returns names of provinces/territories
# us0 <- getData('GADM', country="USA", level=0)
us1 <- getData('GADM', country="USA", level=1)
# us2 <- getData('GADM', country="USA", level=2)
plot(us1)              # state boundaries split at 
                       # the dateline
us1$NAME_1             # returns names of the states + DC
spplot(us1, "ID_1")
spplot(us1, "NAME_1")  # color codes states and
                       # provides their names
#
# Here attempting unsuccessfully to combine U.S. and Canada on one map.
# Attempts at selecting given states or provinces have been unsuccessful.
#
plot(us1,can1)
us.can1 <- rbind(us1,can1)
```
# Try this script 
#To plot multiple SpatialPolygons objects on the same device, one approach is to specify the geographic extent you wish to plot first, and then using plot(..., add=TRUE). This will add to the map only those points that are of interest.

Plotting using a projection, (e.g. a polyconic projection) requires first using the spTransform() function in the rgdal package to make sure all the layers are in the same projection.
```{r}
##Call a new plot
plot.new()
## Specify a geographic extent for the map
## by defining the top-left and bottom-right geographic coordinates
mapExtent <- rbind(c(-74, 59), c(-60, 40))#final set: (c(-74, 59), c(-61, 40))
#plot(mapExtent, col=NA, pch="o", add=TRUE)#trouble-shooting how bit to makethe extent

## Specify the required projection using a proj4 string
## Use http://www.spatialreference.org/ to find the required string
## Polyconic for North America
newProj <- CRS("+proj=poly +lat_0=0 +lon_0=-100 +x_0=0 
            +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs") #this should be same/similar to "+init=epsg:4326" which is what we have for the points above

## Project the map extent (first need to specify that it is longlat) 
mapExtentPr <- spTransform(SpatialPoints(mapExtent, 
                  proj4string=CRS("+proj=longlat")),
                  newProj)
plot(mapExtentPr, pch=NA) #try this to make St Lawrence River
#plot(mapExtentPr,pch="o", add=TRUE) #visualize using points the topleft and bottomright corners of the extent

## Project other layers
can0Pr <- spTransform(can0, newProj)
can1Pr <- spTransform(can1, newProj)
us1Pr <- spTransform(us1, newProj) 

## Plot each projected layer, beginning with the projected extent
#plot(can0Pr, border="black", col=NA, add=TRUE)
#plot(can1Pr, border=NA, col="lightgrey", add=TRUE)
#plot(us1Pr, border="black", col=NA, add=TRUE)

#Adding other features to the map, such as highlighting jurisdictions of interest, can easily be done using the same approach:

## Highlight provinces and states of interest
theseJurisdictions <- c("Québec", "Nova Scotia", "New Brunswick", "Newfoundland and Labrador")
theseUSJurisdictions <- c("Maine")
borderingJurisdictions<- c("Ontario")
borderingUSJurisdictions<- c("New Hampshire", "Vermont", "Massachusetts", "Connecticut", "New York")

p <- plot(can1Pr[can1Pr$NAME_1 %in% theseJurisdictions, ], border="black", 
    col="gray90", add=TRUE) + 
  plot(us1Pr[us1Pr$NAME_1 %in% theseUSJurisdictions, ], border="black", 
    col="gray90", add=TRUE) +
  plot(can1Pr[can1Pr$NAME_1 %in% borderingJurisdictions, ], border="black", cex=0.4, 
    col="white", add=TRUE) +
  plot(us1Pr[us1Pr$NAME_1 %in% borderingUSJurisdictions, ], border="black", cex=0.4, 
    col="white", add=TRUE)

# Try adding points to the map, from our lynx dataset
lynx<-read.csv("home/tl50a/EEMS/WGS_locations_WGS84.csv",header=TRUE)

spdf<-data.frame(samples=lynx$sample,y=lynx$lat,x=lynx$long, class=lynx$pop)
coordinates(spdf)<-c("x","y") #retrieve the spatial coordinates from the lynx dataset
spdfpr<- spTransform(spdf, newProj)
proj4string(spdf)<-CRS("+init=epsg:4326") #sets projection attributes on spatial data #Here we have WGS84
#EPSG:3347	NAD83 / Statistics Canada Lambert	Lambert Conic Conformal (2SP)
spdftrans<-spTransform(spdf, CRS("+proj=poly +lat_0=0 +lon_0=-100 +x_0=0 
            +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs")) #transform to WGS 84

palette(c("orange", "violet", "lightgreen", "tomato", "blueviolet")) #matches our admixture plot
plot(spdftrans,pch=16, col=as.factor(spdftrans$class), add=TRUE) #palette=(rainbow(5))

legend("bottomright", legend=c("N St. Lawrence River", "Newfoundland", "S St. Lawrence River", "Bobcat", "Bobcat-Lynx Hybrid"), pch=16, col=c("orange", "violet", "lightgreen", "tomato", "blueviolet"), text.col="black", cex = 0.6, border=FALSE, box.col=NA, inset=0.1)

#Adding a much larger map for a map "inlay" on top
mapExtent2 <- rbind(c(-156, 80), c(-68, 19))#final set: (c(-74, 59), c(-61, 40)) mapExtent <- rbind(c(-156, 80), c(-68, 19))
#plot(mapExtent, col=NA, pch="o", add=TRUE)#trouble-shooting how bit to makethe extent

## Project the map extent (first need to specify that it is longlat) 
mapExtent2Pr <- spTransform(SpatialPoints(mapExtent, 
                  proj4string=CRS("+proj=longlat")),
                  newProj)
plot(mapExtent2Pr, pch=NA, col="blue")
plot(can0Pr, border="black", col=NA, add=TRUE)

  
  
  
  